### 2018-08-19 paper/finitecomplete-failures.tex
checked the fixes

For a given reference process $P$ and for each integer $p\ge 0$, we define a
CSP test process for failures refinement as shown below.
%
\begin{eqnarray}
U_F(p) & = & U_F(p,\varepsilon)
\\
U_F(p,s) & = & \big( \Extchoice e:(\Sigma - [P/s]^0) @ e \then \efail\then \Stop \big)
\label{eq:ufa}
\\ & & \extchoice \nonumber
\\ & & ([P/s]^0 = \varnothing)    \&   \big( \epass \then \Stop \big)
\label{eq:ufb}
\\ & & \extchoice \nonumber
\\ & & (\#s < p) \& \big( \Extchoice e:[P/s]^0 @ e \then U_F(p,s.e) \big)
\label{eq:ufc}
\\ & & \extchoice \nonumber
\\ & & (\#s = p) \& \big( \sqcap_{H\in\text{minHit}(P/s)} ( \Extchoice e:H @ e \then \epass \then\Stop   )  \big)
\label{eq:ufd}
\end{eqnarray}
%
A test is performed by running $U_F(p)$ concurrently with any SUT process $Q$
\fixme{alcc: add an example?}\fxnote{jp: I think that the following explanations, together with examples in Section~\ref{sec:case}, suffice.} that operates on the same alphabet as $P$,
synchronising over alphabet $\Sigma$. Therefore, a test execution is any
trace of the concurrent process
\[
Q\parallel[\Sigma] U_F(p).
\]
It is assumed that the events $\efail$ and $\epass$, denoting FAIL and PASS
of the test execution, are events outside $\Sigma$. Since we assume that $Q$
is free of livelocks, it is guaranteed that each test execution terminates
after some $s\in\trc(P)$ with length $(p+1)$ at the latest. The test is
\emph{passed} by the SUT (written $Q\ \pass\ U_F(p)$) if, and only if, {\it
every} execution of $Q\parallel[\Sigma] U_F(p)$ terminates with PASS event
$\epass$. This can also be  expressed by means of a failures refinement.
\[
Q\ \pass\ U_F(p) \equiv (\epass\then\Stop) \lessdet_F (Q\parallel[\Sigma] U_F(p)) \hide \Sigma
\]
This type of pass relation is often called \emph{must test}, because every
test execution must end with the $\epass$
event~\cite{Hennessy:1988:ATP:50497}. Note that it is necessary to use the
failures-refinement relation in this condition, and not the trace-refinement
relation:~$(Q\parallel[\Sigma] U_F(p)) \hide \Sigma$ may have  the same
visible traces $\varepsilon$ and $\langle \epass\rangle$ as the ``Test Passed
Process'' $(\epass\then\Stop)$. However, the former may nondeterministically
refuse $\epass$, due to a deadlock occurring when a faulty SUT process
executes concurrently with $U_F(p,s)$ executing branch (\ref{eq:ufd}),
because $\#s = p$. This is explained further in the next paragraphs.

@@

\begin{lemma}\label{lemma:mainfsound}
Test suite $\TS_F$ generated from CSP process $P$, as specified in Theorem~\ref{th:failurestest}, is passed by
every CSP process $Q$ satisfying $P\lessdet_F Q$.
\end{lemma}
\begin{proof}
Suppose that $P\lessdet_F Q$. Then 
$\trc(Q)\subseteq \trc(P)$. Therefore, when running in parallel with $Q$, any
adaptive test $U_F(p)$ will always enter the branches (\ref{eq:ufb}),
(\ref{eq:ufc}),  or (\ref{eq:ufd}) of the external choice construction for
$U_F(p,s)$. Branch~(\ref{eq:ufa}) can never be entered in the prallel execution of $Q$ and
$U_F(p)$, because $[Q/s]^0\subseteq [P/s]^0$ for all traces of $Q$.

Moreover,  Lemma~\ref{lemma:hseta}
implies that for all traces $s\in\trc(Q)$, every $H$ in $\text{minHit}(P/s)$ is
a hitting set for $\minaccs(Q/s)$. Branch (\ref{eq:ufb}) leads always to a PASS verdict, and branch
(\ref{eq:ufc}) to test continuation without a verdict. For the last branch,
we note that any selected minimal hitting set $H\in\text{minHit}(P/s)$ has a
non-empty intersection with each of the minimal acceptances of $Q/s$. As a
consequence, $Q/s$ never blocks when offered events from $H$, and the test
terminates with PASS event $\epass$. Note that this argument requires that
$Q$ is free of livelocks, because otherwise the PASS-events might not become
visible, due to unbounded sequences of hidden events performed by $Q$. 
%Note
%further, that this proof did not refer in any way to the size of $Q$'s
%normalised transition graph, so the test suite is sound for {\it all}
%non-divergent CSP process refining $P$. 
\fixme{alcc: Soundness does not
depend on the fault domain. I would have it as a separate theorem.}
\fxnote{jp: this is a good idea, I have split the proof of the main theorem in 2 lemmas.}
\xbox
\end{proof}

@@

For Case~1, consider a  trace $s.e\in\trc(Q)$ such that
$s\in\trc(P)$, but $s.e\not\in\trc(P)$. Such a trace always exists because
$\varepsilon$ is a trace of every process. In this case, $s$ is also a trace
of the product graph $G = G(P)\times G(Q)$ defined in Section~\ref{sec:ntg}.
Suppose that $G/s = (n_P,n_Q)$.
The length of $s$ is not known, but 
from the construction of $G$,  we know that $G$ has at most
$pq$ reachable states, because $G(P)$ has $p$ states, and $G(Q)$ has at most
$q$ states.  Applying
Lemma~\ref{lemma:reachproduc} implies that $(n_P,n_Q)$ can be reached by a trace
$u\in\trc(G)$ of length $\#u < pq$. Now the construction of the transition
function of $G$ implies that $u$ is also a trace of $P$ and $Q$. Since test
$U_F(pq-1)$ accepts all traces of $P$ up to length $pq-1$, $u$ is also a
trace of this test, and, by construction, $U_F(pq-1)/u = U_F(pq-1,u)$. Since
$s.e\not\in\trc(P)$, $e$ is an element of $\Sigma-[P/u]^0$. Therefore, in at
least one execution, $U_F(pq-1,u)$ executes its first branch (\ref{eq:ufa})
with this event $e$, so that the test fails with event $\efail$. Again, the
assumption of non-divergence of Q is needed for this conclusion. \fixme{alcc:
I didn't see how Lemma 3 is being applied here. Also, given the definition of
the test as a process, I'm not sure why you need to make the argument using
graph product.}
\fxnote{jp: I do hope that this is now clearer from the new structure of the lemmas}

>>>>

For a given reference process $P$ and for each integer $p\ge 0$, we define a
CSP test process for failures refinement as shown below.
%
\begin{eqnarray}
U_F(p) & = & U_F(p,\varepsilon)
\\
U_F(p,s) & = & \big( \Extchoice e:(\Sigma - [P/s]^0) @ e \then \efail\then \Stop \big)
\label{eq:ufa}
\\ & & \extchoice \nonumber
\\ & & ([P/s]^0 = \varnothing)    \&   \big( \epass \then \Stop \big)
\label{eq:ufb}
\\ & & \extchoice \nonumber
\\ & & (\#s < p) \& \big( \Extchoice e:[P/s]^0 @ e \then U_F(p,s.e) \big)
\label{eq:ufc}
\\ & & \extchoice \nonumber
\\ & & (\#s = p) \& \big( \sqcap_{H\in\text{minHit}(P/s)} ( \Extchoice e:H @ e \then \epass \then\Stop   )  \big)
\label{eq:ufd}
\end{eqnarray}
%
A test is performed by running $U_F(p)$ concurrently with any SUT process $Q$
%\fixme{alcc: add an example?}\fxnote{jp: I think that the following explanations, together with examples in Section~\ref{sec:case}, suffice.} that operates on the same alphabet as $P$,
synchronising over alphabet $\Sigma$. Therefore, a test execution is any
trace of the concurrent process
\[
Q\parallel[\Sigma] U_F(p).
\]
It is assumed that the events $\efail$ and $\epass$, denoting FAIL and PASS
of the test execution, are events outside $\Sigma$. Since we assume that $Q$
is free of livelocks, it is guaranteed that each test execution terminates
after some $s\in\trc(P)$ with length $(p+1)$ at the latest. The test is
\emph{passed} by the SUT (written $Q\ \pass\ U_F(p)$) if, and only if, {\it
every} execution of $Q\parallel[\Sigma] U_F(p)$ terminates with PASS event
$\epass$. This can also be  expressed by means of a failures refinement.
\[
Q\ \pass\ U_F(p) \equiv (\epass\then\Stop) \lessdet_F (Q\parallel[\Sigma] U_F(p)) \hide \Sigma
\]
This type of pass relation is often called \emph{must test}, because every
test execution must end with the $\epass$
event~\cite{Hennessy:1988:ATP:50497}. Note that it is necessary to use the
failures-refinement relation in this condition, and not the trace-refinement
relation:~$(Q\parallel[\Sigma] U_F(p)) \hide \Sigma$ may have  the same
visible traces $\varepsilon$ and $\langle \epass\rangle$ as the ``Test Passed
Process'' $(\epass\then\Stop)$. However, the former may nondeterministically
refuse $\epass$, due to a deadlock occurring when a faulty SUT process
executes concurrently with $U_F(p,s)$ executing branch (\ref{eq:ufd}),
because $\#s = p$. This is explained further in the next paragraphs.

@@

\begin{lemma}\label{lemma:mainfsound}
Test suite $\TS_F$ generated from CSP process $P$, as specified in Theorem~\ref{th:failurestest}, is passed by
every CSP process $Q$ satisfying $P\lessdet_F Q$.
\end{lemma}
\begin{proof}
Suppose that $P\lessdet_F Q$. In this case, in particular, $\trc(Q)\subseteq
\trc(P)$. Therefore, when running in parallel with $Q$, any adaptive test
$U_F(p)$ always enters the branches (\ref{eq:ufb}), (\ref{eq:ufc}), or
(\ref{eq:ufd}) of the external choice construction for $U_F(p,s)$.
Branch~(\ref{eq:ufa}) can never be entered in the parallel execution of $Q$
and $U_F(p)$, because $[Q/s]^0\subseteq [P/s]^0$ for all traces of $Q$.

Moreover,  Lemma~\ref{lemma:hseta}
implies that for all traces $s\in\trc(Q)$, every $H$ in $\text{minHit}(P/s)$ is
a hitting set for $\minaccs(Q/s)$. Branch (\ref{eq:ufb}) leads always to a PASS verdict, and branch
(\ref{eq:ufc}) to test continuation without a verdict. For the last branch,
we note that any selected minimal hitting set $H\in\text{minHit}(P/s)$ has a
non-empty intersection with each of the minimal acceptances of $Q/s$. As a
consequence, $Q/s$ never blocks when offered events from $H$, and the test
terminates with PASS event $\epass$. Note that this argument requires that
$Q$ is free of livelocks, because otherwise the PASS-events might not become
visible, due to unbounded sequences of hidden events performed by $Q$.
%Note
%further, that this proof did not refer in any way to the size of $Q$'s
%normalised transition graph, so the test suite is sound for {\it all}
%non-divergent CSP process refining $P$.
%\fixme{alcc: Soundness does not
%depend on the fault domain. I would have it as a separate theorem.}
%\fxnote{jp: this is a good idea, I have split the proof of the main theorem in 2 lemmas.}
\xbox
\end{proof}

@@

For Case~1, consider a  trace $s.e\in\trc(Q)$ such that $s\in\trc(P)$, but
$s.e\not\in\trc(P)$. Such a trace always exists because $\varepsilon$ is a
trace of every process. In this case, $s$ is also a trace of the product
graph $G = G(P)\times G(Q)$ defined in Section~\ref{sec:ntg}. Suppose that
$G/s = (n_P,n_Q)$. The length of $s$ is not known, but from the construction
of $G$,  we know that $G$ has at most $pq$ reachable states, because $G(P)$
has $p$ states, and $G(Q)$ has at most $q$ states. By
Lemma~\ref{lemma:reachproduc}, $(n_P,n_Q)$ can be reached by a trace
$u\in\trc(G)$ of length $\#u < pq$. Now the construction of the transition
function of $G$ implies that $u$ is also a trace of $P$ and $Q$. Since test
$U_F(pq-1)$ accepts all traces of $P$ up to length $pq-1$, $u$ is also a
trace of this test, and, by construction, $U_F(pq-1)/u = U_F(pq-1,u)$. Since
$s.e\not\in\trc(P)$, $e$ is an element of $\Sigma-[P/u]^0$. So, in at least
one execution, $U_F(pq-1,u)$ executes its first branch (\ref{eq:ufa}) with
this event $e$, so that the test fails. Again, the
assumption of non-divergence of Q is needed for this conclusion. %\fixme{alcc:
%I didn't see how Lemma 3 is being applied here. Also, given the definition of
%the test as a process, I'm not sure why you need to make the argument using
%graph product.}
%\fxnote{jp: I do hope that this is now clearer from the new structure of the lemmas}

### 2018-08-19 paper/introduction.tex
checked the fixes

This paper presents complete black-box test suites for software and systems modelled using the {\it Communicating Sequential Processes (CSP)} process 
algebra~\cite{Hoare:1985:CSP:3921,Roscoe:1997:TPC:550448}. This process algebra has been widely used for the specification and verification of communicating 
concurrent control systems, many of these applications have been described in the monograph~\cite{Roscoe2010} and the references 
contained there.

The complete testing theory presented here applies to 
CSP processes that are
divergence-free\footnote{The assumption of divergence freedom is usually
applied in black-box testing, since it cannot distinguish between divergence
and deadlock.} and interpreted both in the trace and the failures semantics.
Our results complement work published 
in~\cite{DBLP:conf/pts/CavalcantiS17}. There, fault domains are specified as
collections of processes refining a  ``most general'' fault domain member.
With that concept, complete test suites may be finite or infinite. This gives
important insight into the theory of fault-domain testing for CSP, but we are
particularly interested in {\it finite} suites when it comes to practical
application. While~\cite{DBLP:conf/pts/CavalcantiS17} requires additional
criteria to select tests from infinite test suites, here, we further restrict
fault domains using a graph representation of processes used in model
checking to obtain test suites that are finite.

Our complementary approach to the definition of CSP fault domains is
presented in this paper. We observe that every finite-state CSP process can
be semantically represented as a finite normalised transition graph, whose
edges are labelled by the events the process engages in, and whose nodes are
labelled by minimal acceptances or, alternatively, maximal
refusals~\cite{Roscoe:1994:CME:197600}. The maximal refusals express the
degree of nondeterminism present in a given process state that is in
one-one-correspondence to a node of the normalised transition graph. Inspired
by the way that fault-domains are specified for finite state machines (FSMs), we
define them as the set of CSP processes whose normalised transition graphs do
not exceed the size of the reference model's graph by more than a given
constant.

The main contribution of this paper is the proof that for fault domains of
the described type, finite, complete test suite generation methods can be
given for testing against trace and failures refinement and equivalence. The
existence of -- possibly infinite -- complete test suites has been
established for process algebras, and for CSP in particular, by several
authors~\cite{Hennessy:1988:ATP:50497,Schneider:1995:OST:203471.203475,DBLP:conf/fm/PeleskaS96,peleska1997a,DBLP:conf/pts/CavalcantiS17}.
To the best of our knowledge, however, this article is the first to present
{\it finite}, complete test suites associated with this class of fault
domains and conformance relations. Moreover, our result is not a simple transcription of
existing knowledge  about finite, complete test suites for finite state machines:
the capabilities of CSP to express nondeterminism in a more fine-grained way than it is possible for FSMs requires a more complex approach to testing systems for conformity in the failures-model than it is required for model-based testing against nondeterministic FSMs, as published, for example, in~\cite{hierons_testing_2004,DBLP:conf/hase/PetrenkoY14}.

>>>>

This paper presents complete black-box test suites for software and systems
modelled using the {\it Communicating Sequential Processes (CSP)} process
algebra~\cite{Hoare:1985:CSP:3921,Roscoe:1997:TPC:550448}. This process
algebra has been widely used for the specification and verification of
communicating concurrent control systems. Many of these applications have
been described in~\cite{Roscoe2010} and in the references there.

The complete testing theory presented here applies to CSP processes that are
divergence-free\footnote{The assumption of divergence freedom is usually
applied in black-box testing, since it cannot distinguish between divergence
and deadlock.} and interpreted both in the trace and the failures semantics.
Our results complement work published in~\cite{DBLP:conf/pts/CavalcantiS17}.
There, fault domains are specified as collections of processes refining a
``most general'' fault domain member. With that concept, complete test suites
may be finite or infinite. This gives important insight into the theory of
fault-domain testing for CSP, but we are particularly interested in {\it
finite} suites when it comes to practical application.
While~\cite{DBLP:conf/pts/CavalcantiS17} may require additional criteria to
select tests from still infinite test suites, here, we further restrict fault
domains using a graph representation of processes used in model checking to
obtain test suites that are finite.

Our complementary approach to the definition of CSP fault domains is
presented in this paper. We observe that every finite-state CSP process can
be represented as a finite normalised transition graph, whose edges are
labelled by the events the process engages in, and whose nodes are labelled
by minimal acceptances or, alternatively, maximal
refusals~\cite{Roscoe:1994:CME:197600}. The maximal refusals express the
degree of nondeterminism present in a given process state that is in
one-one-correspondence to a node of the normalised transition graph. Inspired
by the way that fault-domains are specified for finite state machines (FSMs),
we define them as the set of CSP processes whose normalised transition graphs
do not exceed the size of the reference model's graph by more than a given
constant.

The main contribution of this paper is the proof that for fault domains of
the described type, finite, complete test suite generation methods can be
given for testing against trace and failures refinement and equivalence. The
existence of -- possibly infinite -- complete test suites has been
established for process algebras, and for CSP in particular, by several
authors~\cite{Hennessy:1988:ATP:50497,Schneider:1995:OST:203471.203475,DBLP:conf/fm/PeleskaS96,peleska1997a,DBLP:conf/pts/CavalcantiS17}.
To the best of our knowledge, however, this article is the first to present
{\it finite}, complete test suites associated with this class of fault
domains and conformance relations.

Moreover, our result is not a simple transcription of existing knowledge
about finite, complete test suites for finite state machines: the
capabilities of CSP to express nondeterminism in a more fine-grained way than
it is possible for FSMs requires a more complex approach to testing systems
for conformity in the failures-model than it is required for model-based
testing against nondeterministic FSMs, as published, for example,
in~\cite{hierons_testing_2004,DBLP:conf/hase/PetrenkoY14}.
